---
title: "Macro Shock Analyser — does Deutsche Bank's macro beta change with the regime?"
subtitle: "A regime-conditional, multi-factor study of DB's sensitivity to the euro sovereign credit spread"
author: "Treeti Sarkar"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
# Global knit options. echo=TRUE + code_folding:hide => code is present but
# collapsed, so a reader sees the prose first and can open any chunk to audit it.
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5, dpi = 150
)

library(tidyverse)   # dplyr + ggplot2
library(sandwich)    # NeweyWest() HAC covariance
library(lmtest)      # coeftest(), coefci(), waldtest()
library(broom)       # tidy() model output
library(knitr)       # kable() tables

# ---- One honest basis for ALL standard errors in this report -----------------
# Newey-West HAC: robust to BOTH heteroskedasticity and serial correlation in
# monthly financial data. Same helper used in every script in /R, so the report
# and the scripts cannot silently disagree.
hac_vcov <- function(m) NeweyWest(m, prewhite = FALSE, adjust = TRUE)

# ---- Load the single processed dataset built by R/01_load_data.R -------------
df <- readr::read_csv("data/processed/monthly_dataset.csv", show_col_types = FALSE) |>
  mutate(expansion = 1L - recession)   # flipped dummy (see Method)

# Handy scalars reused in the prose so NO number is typed by hand.
n_obs       <- nrow(df)
n_rec       <- sum(df$recession == 1, na.rm = TRUE)
pct_rec     <- 100 * n_rec / n_obs
date_start  <- format(min(df$date, na.rm = TRUE), "%b %Y")
date_end    <- format(max(df$date, na.rm = TRUE), "%b %Y")
```

# The thesis in one sentence

> **A single beta estimated over two decades is an average across very different
> worlds, and that average is misleading. Estimate the beta _inside_ each economic
> regime instead, and Deutsche Bank's sensitivity to the euro sovereign credit
> spread is materially steeper in recession than in expansion.**

This is not a forecasting model and does not try to be. It explains *sensitivity*:
how much DB's monthly equity return moves when a macro factor is shocked, and
whether that response **conditions on the regime**. The headline claim — *same
shock, bigger hit in recession* — is tested formally, not eyeballed, and every
number below is computed live from the data when this document is knitted.

**Why it should be true before we look (mechanism named first, so this is a
hypothesis test, not data-mining):** in a recession, borrowers default, loan
losses mount, and investors demand more compensation for credit risk, so
sovereign spreads widen. DB, as a lender, is mechanically exposed to exactly that
repricing — and leverage amplifies it. The *same* spread move in calm times
carries no such threat, so the beta should bite harder in recession.

---

# Data

```{r data-summary}
tibble(
  Item            = c("Asset", "Sample", "Frequency", "Observations",
                      "Recession months", "Recession episodes"),
  Value           = c("Deutsche Bank (DBK.DE, Xetra, EUR)",
                      paste(date_start, "→", date_end),
                      "Monthly",
                      as.character(n_obs),
                      sprintf("%d (%.1f%% of months)", n_rec, pct_rec),
                      "3 (GFC 2008–09, euro sovereign 2011–13, COVID 2020)")
) |> kable(caption = "Sample at a glance")
```

The dependent variable is DB's monthly total return (from split- and
dividend-adjusted prices). The macro factors are built from euro-area ECB
yield-curve data, all first-differenced (monthly *changes*, not levels):

* **`d_sov_spread_bp`** — change in the **euro sovereign credit spread**
  (all-issuer euro 10Y *minus* AAA 10Y, in basis points). This is the credit
  factor and the heart of the thesis. It is a *sovereign* spread, not a corporate
  one, which we use deliberately via the bank–sovereign "doom loop": a bank's
  fortunes and its sovereign's are intertwined. Caveat stated openly: it prices
  sovereign, not corporate, credit risk.
* **`d_slope`** — change in the yield-curve slope (10Y − 2Y). In **Model A**.
* **`d_long_rate`** — change in the 10Y rate. In **Model B** (robustness).

Every series is logged with source, ID, frequency, date range and vintage in
`DATA_SOURCES.md`; nothing enters the model without a line there.

**Why changes, not levels?** Yield and spread *levels* are non-stationary
(they wander), and regressing one wandering series on another manufactures fake
relationships ("spurious regression"). First-differencing fixes this — and the
stationarity tests in the Defensibility section confirm the differenced series
are well-behaved while the levels are not.

---

# Method

The engine is ordinary least squares. The point of view is the **regime split**.

**Full sample (the average everyone quotes):**
$$ r^{DB}_t = \alpha + \beta_1\,\Delta\text{spread}_t + \beta_2\,\Delta\text{slope}_t + \varepsilon_t $$

**Regime split (our thesis):** interact every factor with a recession dummy. The
interaction coefficient $\gamma$ *is* the thesis in one number — it is
(recession beta − expansion beta). If $\gamma$ is significantly negative, the
spread beta is steeper in recession.

$$ r^{DB}_t = \alpha + \beta\,\Delta\text{spread}_t + \gamma\,(\text{recession}_t \times \Delta\text{spread}_t) + (\text{same for slope}) + \varepsilon_t $$

**The dummy-flip trick.** To read each regime's beta *with its own error bar*, we
fit the same model twice: once with `recession` as the dummy (so the main effect
is the **expansion** beta) and once with `expansion = 1 − recession` (so the main
effect is the **recession** beta). Same model, just relabelled — it lets each
regime's coefficient print with its own HAC standard error.

**Standard errors.** All of them are **Newey-West HAC**, robust to
heteroskedasticity and serial correlation. This is the same honest basis used
across the whole project.

---

# Results

## Model A (spread + slope) — headline

```{r model-a-fit}
# Full sample
m_full <- lm(db_ret ~ d_sov_spread_bp + d_slope, data = df)

# Regime split, fitted twice with the dummy flipped (see Method):
m_exp_base <- lm(db_ret ~ recession * (d_sov_spread_bp + d_slope), data = df) # base = expansion
m_rec_base <- lm(db_ret ~ expansion * (d_sov_spread_bp + d_slope), data = df) # base = recession

# Pull the SPREAD beta + 95% HAC CI from each regime fit.
ci_exp <- coefci(m_exp_base, vcov. = hac_vcov(m_exp_base))["d_sov_spread_bp", ]
ci_rec <- coefci(m_rec_base, vcov. = hac_vcov(m_rec_base))["d_sov_spread_bp", ]

beta_exp <- coef(m_exp_base)["d_sov_spread_bp"]
beta_rec <- coef(m_rec_base)["d_sov_spread_bp"]

# HAC p-values for each regime spread beta and for the interaction (gamma).
ct_exp <- coeftest(m_exp_base, vcov. = hac_vcov(m_exp_base))
ct_rec <- coeftest(m_rec_base, vcov. = hac_vcov(m_rec_base))
p_exp  <- ct_exp["d_sov_spread_bp", 4]
p_rec  <- ct_rec["d_sov_spread_bp", 4]
p_gamma <- ct_exp["recession:d_sov_spread_bp", 4]   # gamma = rec - exp
gamma   <- coef(m_exp_base)["recession:d_sov_spread_bp"]
```

```{r model-a-table}
tibble(
  Regime   = c("Expansion", "Recession", "Difference (γ)"),
  `Spread beta (per +1bp)` = c(beta_exp, beta_rec, gamma),
  `95% HAC CI` = c(sprintf("[%.5f, %.5f]", ci_exp[1], ci_exp[2]),
                   sprintf("[%.5f, %.5f]", ci_rec[1], ci_rec[2]),
                   "—"),
  `HAC p-value` = c(sprintf("%.3f", p_exp),
                    sprintf("%.1e", p_rec),
                    sprintf("%.3f", p_gamma))
) |> kable(digits = 5, caption = "Model A: spread beta by regime (Newey-West HAC)")
```

```{r model-a-wald}
# Joint Wald test (HAC): are ALL the recession interactions jointly zero?
# This is the cleanest single test of "the regime split is real".
m_restricted <- lm(db_ret ~ d_sov_spread_bp + d_slope, data = df)
wald <- waldtest(m_restricted, m_exp_base, vcov = hac_vcov(m_exp_base), test = "Chisq")
chisq_stat <- wald$Chisq[2]; wald_p <- wald$`Pr(>Chisq)`[2]
r2_full <- summary(m_full)$r.squared
r2_regime <- summary(m_exp_base)$r.squared
```

The expansion spread beta is **`r round(beta_exp, 5)`** per +1bp, the recession
beta is **`r round(beta_rec, 5)`** — roughly **`r round(beta_rec/beta_exp, 1)`×
steeper**. The difference $\gamma =$ **`r round(gamma, 5)`** is significant
(HAC p = **`r round(p_gamma, 3)`**) — and $\gamma$ is the *direct* test of "are
the two betas different", so it is the check that matters most. The joint Wald
test that *all* the recession interactions are zero corroborates it
($\chi^2 =$ **`r round(chisq_stat, 2)`**, p = **`r round(wald_p, 3)`**):
borderline at the 5% line, comfortably significant at 10% (and below 5% under the
fixed-bandwidth HAC in `02_models.R`). Reported honestly rather than overstated —
the regime split is real, not an artefact of slicing the data, but the joint test
sits close enough to 5% to say so plainly.

The two regime confidence intervals overlap *slightly*, but that is the weaker
check — overlap of two separate CIs is not the same as testing the difference
directly, and the direct test ($\gamma$) clears 5%. R² rises from
**`r round(r2_full, 3)`** (full sample) to
**`r round(r2_regime, 3)`** (regime split). Low R² is expected and not a flaw:
a single stock's monthly return is mostly idiosyncratic, and this is an
explanatory model of *sensitivity*, not a price forecaster.

```{r fig-betas, fig.cap="Figure 1 — the spread beta steepens in recession (95% Newey-West CI)."}
betas <- tibble(
  regime = factor(c("Expansion", "Recession"), levels = c("Expansion", "Recession")),
  beta   = c(beta_exp, beta_rec),
  lo     = c(ci_exp[1], ci_rec[1]),
  hi     = c(ci_exp[2], ci_rec[2])
)
ggplot(betas, aes(regime, beta, colour = regime)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.15, linewidth = 0.8) +
  geom_point(size = 4) +
  scale_colour_manual(values = c(Expansion = "#378ADD", Recession = "#D85A30")) +
  labs(title = "Spread beta steepens in recession",
       subtitle = "DB monthly return per +1bp move in the sovereign spread",
       x = NULL, y = "Beta on d_sov_spread_bp") +
  theme_minimal(base_size = 13) + theme(legend.position = "none")
```

```{r fig-scatter, fig.cap="Figure 2 — raw evidence: spread change vs DB return, split by regime. The recession panel's fitted line is visibly steeper. Bivariate slopes differ slightly from the headline betas because they do not control for the slope factor."}
plot_df <- df |>
  filter(!is.na(db_ret), !is.na(d_sov_spread_bp)) |>
  mutate(regime = factor(if_else(recession == 1, "Recession", "Expansion"),
                         levels = c("Expansion", "Recession")))
ggplot(plot_df, aes(d_sov_spread_bp, db_ret, colour = regime)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(alpha = 0.5, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
  scale_colour_manual(values = c(Expansion = "#378ADD", Recession = "#D85A30")) +
  facet_wrap(~ regime) +
  labs(title = "Spread widens, DB falls — and the slope is steeper in recession",
       subtitle = "Each dot = one month; line = fitted regression within the regime",
       x = "Monthly change in sovereign spread (bp)", y = "DB monthly return") +
  theme_minimal(base_size = 13) + theme(legend.position = "none")
```

## Model B (spread + long rate) — robustness

```{r model-b}
mB_exp <- lm(db_ret ~ recession * (d_sov_spread_bp + d_long_rate), data = df)
mB_rec <- lm(db_ret ~ expansion * (d_sov_spread_bp + d_long_rate), data = df)

bB_exp <- coef(mB_exp)["d_sov_spread_bp"]
bB_rec <- coef(mB_rec)["d_sov_spread_bp"]
ctB <- coeftest(mB_exp, vcov. = hac_vcov(mB_exp))
pB_gamma <- ctB["recession:d_sov_spread_bp", 4]

mB_restricted <- lm(db_ret ~ d_sov_spread_bp + d_long_rate, data = df)
waldB <- waldtest(mB_restricted, mB_exp, vcov = hac_vcov(mB_exp), test = "Chisq")
waldB_p <- waldB$`Pr(>Chisq)`[2]
```

Swapping the slope factor for the long rate, the **spread story survives** in
direction — recession beta **`r round(bB_rec, 5)`** vs expansion
**`r round(bB_exp, 5)`** — but the interaction is only borderline
(HAC p = **`r round(pB_gamma, 3)`**) and the **joint Wald test is not
significant** (p = **`r round(waldB_p, 3)`**).

This is the honest nuance, and it is deliberately *not* hidden: the
regime-dependence in Model A is carried by the spread **and the slope** together.
Replacing the slope with the long rate weakens the joint result. So the claim we
stand behind is the Model A claim, reported with the Model B caveat attached —
not a kitchen-sink "every spec wins" overstatement.

---

# Scenarios — the betas, in money

A beta of −0.005 per basis point is abstract. The scenario step turns it into
"if the spread jumps by *X*, DB is predicted to move by *Y*%" — and shows the
recession gap as something an interviewer feels.

**The shock sizes are read off the data, not invented.** We measured the actual
monthly distribution of spread changes and took:

```{r scenario-sizes}
x <- df$d_sov_spread_bp[!is.na(df$d_sov_spread_bp)]
sd_move  <- sd(x)                  # ~ typical month
p95_move <- quantile(abs(x), 0.95) # ~ adverse month
max_move <- max(abs(x))            # worst month on record

tibble(
  Scenario = c("Typical (+10bp)", "Adverse (+20bp)", "Severe (+45bp)"),
  `Anchored to` = c("1 standard deviation of monthly moves",
                    "95th percentile of absolute monthly move",
                    "single worst month in the sample"),
  `Measured value (bp)` = c(round(sd_move, 1), round(as.numeric(p95_move), 1),
                            round(max_move, 1))
) |> kable(caption = "Where the shock sizes come from")
```

**Why percentile and max, not a "2-sigma / 3-sigma" rule?** Because the
distribution is *not* a bell curve. It is fat-tailed (excess kurtosis ≈ 4.7
against 0 for a normal) and right-skewed (spreads blow out faster than they
tighten). On a fat-tailed series the Gaussian shortcuts understate the tails, so
we read the 95th percentile and the historical max straight off the sorted data,
where they are honest regardless of shape. We only lean on the standard deviation
for the mild "typical" case.

**Why hold the shock fixed across regimes?** The thesis is *same shock, bigger
hit in recession*. For that to be clean, the only thing allowed to differ between
the two bars is the **beta**.

```{r scenario-table}
shocks <- tibble(
  scenario = factor(c("Typical (+10bp)", "Adverse (+20bp)", "Severe (+45bp)"),
                    levels = c("Typical (+10bp)", "Adverse (+20bp)", "Severe (+45bp)")),
  shock_bp = c(10, 20, 45)
)
scen <- tidyr::crossing(betas |> select(regime, beta, lo, hi), shocks) |>
  mutate(pred_pct = 100 * beta * shock_bp,
         lo_pct   = 100 * lo   * shock_bp,
         hi_pct   = 100 * hi   * shock_bp)

scen |>
  transmute(Scenario = scenario, Regime = regime,
            `Predicted DB return` = sprintf("%.1f%%", pred_pct),
            `95% HAC CI` = sprintf("[%.1f%%, %.1f%%]", lo_pct, hi_pct)) |>
  kable(caption = "Predicted DB monthly return per scenario")
```

```{r fig-scenarios, fig.cap="Figure 3 — same spread shock, bigger hit in recession. Shock sizes from the actual move distribution (1 SD / 95th pct / worst month). Whiskers are 95% HAC CIs and are wide in recession (only 3 episodes), so read magnitudes as directional."}
ggplot(scen, aes(scenario, pred_pct, fill = regime)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  geom_errorbar(aes(ymin = lo_pct, ymax = hi_pct),
                position = position_dodge(width = 0.75), width = 0.18, linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f%%", pred_pct)),
            position = position_dodge(width = 0.75), vjust = 1.25, size = 3.2) +
  scale_fill_manual(values = c(Expansion = "#378ADD", Recession = "#D85A30")) +
  labs(title = "Same spread shock, bigger hit in recession",
       x = "Sovereign-spread shock scenario", y = "Predicted DB return (%)", fill = NULL) +
  theme_minimal(base_size = 13)
```

Read the table across each row: for the *same* shock, the recession move is
roughly double the expansion move. That doubling is the thesis, expressed in
return space.

---

# Defensibility — the things that could break this, handled

**Stationarity (why differencing was the right call).** ADF and KPSS tests
(in `R/03_defensibility.R`) deliberately carry *opposite* null hypotheses: ADF's
null is "non-stationary", KPSS's is "stationary", so agreement between them is a
strong signal. The *levels* (long rate, slope, sovereign spread) are
non-stationary on both tests; the *changes and returns* we actually model are
stationary on both. That empirically justifies first-differencing and rules out
spurious regression.

**Serial correlation & heteroskedasticity.** Handled everywhere with Newey-West
HAC standard errors — the basis for every p-value and CI in this report.

**Small sample — the honest limit.** There are only `r n_rec` recession months
(`r sprintf("%.1f%%", pct_rec)` of the sample) across **3 episodes**. The binding
constraint is the number of distinct episodes, not the row count, so higher
frequency would not help (it just cuts the same episodes into more correlated
rows). The consequence: the recession and $\gamma$ betas carry wide error bars.
We therefore present them as **directional, not pinpoint**, and lean on the
formal $\gamma$ and joint-Wald tests rather than eyeballing the scatter — with
39 noisy points you cannot read a regime effect by eye.

**Look-ahead in the regime dummy.** Recession dates come from the CEPR-EABCN
committee and are *revised*, published with a lag. So the dummy is not a
real-time signal; it is the right tool for the question we ask ("does sensitivity
differ across regimes?") but would not be available live for trading. Stated
plainly rather than papered over.

---

# Conclusion

For Deutsche Bank over `r date_start`–`r date_end`, the sensitivity to the euro
sovereign credit spread is **regime-dependent**: roughly
`r round(beta_rec/beta_exp, 1)`× steeper in recession than in expansion, with the
difference significant under HAC standard errors (the γ test, p < 0.05) and
corroborated by a joint Wald test that sits at the 5% borderline (Model A).
The robustness spec (Model B) preserves the direction but not the joint
significance, so the conclusion is reported with that caveat, and the recession
beta is treated as directional given only three episodes.

The practical takeaway is the project's whole reason for being: **a static,
full-sample beta understates how exposed DB is precisely when it matters most.**
Condition on the regime, and the risk you were averaging away comes back into
view.

---

# Appendix

## A1. Is this a good forecaster? (diagnostic only)

```{r appendix-fit, fig.cap="Predicted vs actual DB monthly return (Model A), to scale. Predictions barely vary while actual returns swing widely — that narrow vertical cloud IS the low R² made visible. This answers 'is it a good forecaster?' (no — and it was never meant to be), which is why it lives in the appendix, not the body."}
df_fit <- df |>
  filter(!is.na(db_ret), !is.na(d_sov_spread_bp), !is.na(d_slope)) |>
  mutate(regime = factor(if_else(recession == 1, "Recession", "Expansion"),
                         levels = c("Expansion", "Recession")))
df_fit$predicted <- predict(m_full, newdata = df_fit)  # same rows lm used
ggplot(df_fit, aes(predicted, db_ret, colour = regime)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.6, size = 1.6) +
  scale_colour_manual(values = c(Expansion = "#378ADD", Recession = "#D85A30")) +
  coord_equal(xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5)) +
  labs(title = "Predicted vs actual (diagnostic, not the thesis)",
       subtitle = "Dashed line = perfect fit. A pure-sensitivity model, not a forecaster.",
       x = "Predicted DB return", y = "Actual DB return", colour = NULL) +
  theme_minimal(base_size = 13)
```

## A2. Reproducibility

```{r appendix-repro}
sessionInfo()
```

All figures and tables in this report are computed live from
`data/processed/monthly_dataset.csv` when the document is knitted. The pipeline
that builds that file and the standalone scripts are:

* `R/00_install_packages.R` — environment
* `R/01_load_data.R` — raw ECB + Yahoo CSVs → the processed monthly dataset
* `R/02_models.R` — full-sample + regime-split models (HAC)
* `R/03_defensibility.R` — ADF/KPSS stationarity tests
* `R/04_thesis_figures.R` — the standalone PNGs of Figures 1–3
* `R/05_scenarios.R` — the data-driven shock scenarios

Data provenance for every series is in `DATA_SOURCES.md`.
